---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = file.path("man", "figures", "README-"),
  out.width = "100%"
)
```

<!-- badges: start -->
[![Project Status: Concept – Minimal or no implementation has been done yet, or the repository is only intended to be a limited example, demo, or proof-of-concept.](https://www.repostatus.org/badges/latest/concept.svg)](https://www.repostatus.org/#concept)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License](https://img.shields.io/badge/license-GPL--3-blue.svg?style=flat)](https://www.gnu.org/licenses/gpl-3.0.html)
[![Release](https://img.shields.io/github/release/inbo/waterbirds1.1.svg)](https://github.com/inbo/waterbirds1.1/releases)
![GitHub](https://img.shields.io/github/license/inbo/waterbirds1.1)
[![R build status](https://github.com/inbo/waterbirds1.1/workflows/check%20package%20on%20main/badge.svg)](https://github.com/inbo/waterbirds1.1/actions)
![r-universe name](https://inbo.r-universe.dev/badges/:name?color=c04384)
![r-universe package](https://inbo.r-universe.dev/badges/waterbirds1.1)
[![Codecov test coverage](https://codecov.io/gh/inbo/waterbirds1.1/branch/main/graph/badge.svg)](https://app.codecov.io/gh/inbo/waterbirds1.1?branch=main)
![GitHub code size in bytes](https://img.shields.io/github/languages/code-size/inbo/waterbirds1.1.svg)
![GitHub repo size](https://img.shields.io/github/repo-size/inbo/waterbirds1.1.svg)
<!-- badges: end -->

# waterbirds1.1

The goal of waterbirds1.1 is to calculate the nutrient input by waterbirds, based on the tool Waterbirds v1.1 (2007) from S. Bauwer & S. Hahn, Netherlands Institute of Ecology (NIOO-KNAW) and the related publications of _Hahn et al. (2007 and 2008)_.

References:

Hahn S., Bauwer S. & Klaassen M. (2007). Estimating the contribution of carnivorous waterbirds to nutrient loading in freshwater habitats. Freshwater Biology 52: 2421-2433. [doi:10.1111/j.1365-2427.2007.01838.x](https://doi.org/10.1111/j.1365-2427.2007.01838.x)

Hahn S., Bauwer S. & Klaassen M. (2008). Quantification of allochthonous nutrient input into freshwater bodies by herbivorous waterbirds. Freshwater Biology 53: 181-193. [doi:10.1111/j.1365-2427.2007.01881.x](https://doi.org/10.1111/j.1365-2427.2007.01881.x)

## Installation

You can install the package from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("inbo/waterbirds1.1")
```

## Calculate nutrient input with `waterbirds1.1`

_Hahn et al. (2007 and 2008)_ presented 2 different models for calculating the nutrient input of nitrogen and phosphorus by waterbirds:

- an intake model that calculates the nutrient loadings based on daily food intake by waterbirds

- a dropping (_Hahn et al., 2008_) or excretion model (_Hahn et al., 2007_) based on the daily faecal output

R-package `waterbirds1.1` has generic functions `give_intake_model()` and `give_dropping_model()` for calculating these models that require a dataset with abundance data (and some predefined variables that default to the values suggested in _Hahn et al. (2007 and 2008)_ but that can be customised).
These functions use predefined info to distinguish between carnivores and herbivores, and food type defaults to grass in case of the intake model of herbivores.

A minimal code example on how to use these functions:

```{r examples_generid}
library(waterbirds1.1)

# make (or load) a dataset with abundance data:
dataset <- data.frame(
  species = c("Anas crecca", "Ardea cinerea"),
  month = 3,
  year = 2026,
  n_individuals = 1,
  location = "ZwartWater"
)

give_intake_model(bird_abundance = dataset)

give_dropping_model(bird_abundance = dataset)
```

These generic functions use some more detailed functions in the background that are specific for carnivores or herbivores.
For carnivores, there are additional functions for breeding birds, corresponding to the calculation described in _Hahn et al. (2007)_.
These functions expect the variables in a different format, but they can be used as well.

All documentation on the use of the functions, including an example, can be accessed via `Reference` in the navigation bar.

As there are some caveats in the use of abundance data from [observations.be](https://observations.be/), we elaborate on this in `vignette("observations", package = "waterbirds1.1")`.
